---
title: "Lab 1 Quarto"
author: "James Lawlor"
format: html
editor: visual
---

## Part 1

My script displays a plot that shows life expectancy for each continent, mapped against GDP per capita. It is displayed as a scatter plot, with the continents being differentiated by the colour of the points.

## Code

Here is the code used to display this plot:

```{r}
library(tidyverse)
library(gapminder)

ggplot(gapminder,
       aes(x=lifeExp,
           y=gdpPercap,
           color=continent,
           size = 3)) + geom_point()
```

<br>

## Part 2

For the second part, the script will compare GDP across continents on average in 2007.

## Code

Here is the script used to accomplish this comparison.

```{r}
library(tidyverse)
library(gapminder)

gap_2007 <- filter(gapminder, year==2007)

continentgdp <- gap_2007 %>% group_by(continent) %>% summarise(mean_gdp = mean(gdpPercap))

ggplot(continentgdp,
       aes(x=continent,
           y=mean_gdp,
           fill = continent)) + geom_col() + labs(title = "Average GDP per Capita by Continent 2007",
                                                  x = "Continent",
                                                  y = "Average GDP per Capita")
```

<br>

## Part 3

For the third part, the script will compare median GDP instead.

## Code

Here is the script used to accomplish this comparison.

```{r}
library(tidyverse)
library(gapminder)

gap_2007 <- filter(gapminder, year==2007)

continentgdp <- gap_2007 %>% group_by(continent) %>% summarise(median_gdp = median(gdpPercap))

ggplot(continentgdp,
       aes(x=continent,
           y=median_gdp,
           fill = continent)) + geom_col() + labs(title = "Average GDP per Capita by Continent 2007",
                                                  x = "Continent",
                                                  y = "Average GDP per Capita")
```

<br>

## Part 4

For the fourth part, the script will get the top 10 countries for GDP per continent in 2007.

## Code

Here is the script used to accomplish this comparison.

```{r}
library(tidyverse)
library(gapminder)

gap_2007 <- filter(gapminder, year==2007)

top10 <- gap_2007 %>% group_by(continent) %>% slice_max(gdpPercap, n=10)

ggplot(top10,
       aes(x=reorder(country, gdpPercap),
           y=gdpPercap,
           fill=continent)) + geom_col(show.legend = FALSE) +
  facet_wrap(~ continent, scales="free_y") + coord_flip() +
  labs(title = "Top 10 Countries for each Continent by GDP per Capita, 2007",
       x = "Country",
       y = "GDP per Capita") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))


```

The newer version of the code which includes "scales='free_y'" is much easier to read, as it allows for the plot to take up greater space on the Y access, allowing the elements more room to breathe.
